---
title: "Natural Experiments"
author: "Abishan Arumugavel and Daniel Beeler"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of variables correlating with the home win percentage in the English Premier League

## Introduction
In this analysis, we aim to investigate the factors that correlate with the home win percentage in the English Premier League. We will explore various variables such as attendance, goal difference, and win/loss streaks to understand their impact on the likelihood of a home team winning a match. By analyzing historical data from the Premier League, we can gain insights into the key drivers of home team success and identify patterns that may influence match outcomes.

## Importing Data
Here we import the relevant dataset into our project.
```{r}
library(readr)
library(readxl)

# Premier League data from 2006 to 2018
data <- read.csv("results.csv")

# Attendance rates from 2006 to 2018
attendance <- read_excel("Attendance.xlsx")

# Premier League data of the 2024 season
# We us additionaly this dataset, because this data set has more information concerning each match of the season.
premier_league_2024 = read.csv("premier league.csv")
```

## Data Cleaning
Here we clean the data, and add relevant columns, which we need for our further analysis.
```{r}
# Calculate goal difference
data$goal_difference <- data$home_goals - data$away_goals

# Calculate points for the home team (3 points for win, 1 point for a draw, 0 points for a loss)
data$points_home <- ifelse(data$result == "H", 3, ifelse(data$result == "D", 1, 0))

# Calculate points for the away team (3 points for win, 1 point for a draw, 0 points for a loss)
data$points_away <- ifelse(data$result == "A", 3, ifelse(data$result == "D", 1, 0))

library(dplyr)

attendance <- attendance %>%
  rename(
    season = Saison,
    team = Team,
    average_attendance = `Average Attendance`
  )

data <- data %>%
  left_join(attendance, 
            by = c("season" = "season", 
                   "home_team" = "team")) %>%
  rename(home_average_attendance = average_attendance)

```


## Analysis
### Analysis of win and loss streaks on the wining probability of a team
We have decided to conduct an analysis on the Premier League data for the 2024 season. We will be looking at the performance of each teams and their opponents in terms of win and loss streaks. This analysis will help us understand the momentum of the teams and how it might affect the outcome of future matches. We will also investigate the impact of the venue (home or away) on the team's performance.

Here for we used the data set only from the season 2024, because this was the only dataset we found with the necessary information.
```{r}
# Sort the data by date
premier_league <- premier_league_2024[order(premier_league_2024$Date),]
```


```{r}
# Load required libraries
library(tidyr)
library(stats)
library(ggplot2)
library(gridExtra)

# Create team results dataframe
team_results <- premier_league %>%
  select(Date, Team, Opponent, Result, Venue) %>%
  mutate(Date = as.Date(Date))

# Create opponent results dataframe with flipped outcomes
# Here we need to also take in account that when one team wins, the other team loses and vice versa.
opponent_results <- premier_league %>%
  select(Date, Team, Opponent, Result, Venue) %>%
  mutate(
    Result = case_when(
      Result == "W" ~ "L",
      Result == "L" ~ "W",
      Result == "D" ~ "D"
    ),
    temp = Team,
    Team = Opponent,
    Opponent = temp,
    # The same goes for the venue, if one team plays at home, the other team plays away.
    Venue = ifelse(Venue == "Home", "Away", "Home"),
    Date = as.Date(Date)
  ) %>%
  select(-temp)

# Combine and calculate streaks
all_results <- rbind(team_results, opponent_results) %>%
  arrange(Date, Team) %>%
  group_by(Team) %>%
  arrange(Date) %>%
  mutate(
    win_streak = cumsum(Result == "W") - cumsum(Result != "W" & lag(Result == "W", default = FALSE)),
    lose_streak = cumsum(Result == "L") - cumsum(Result != "L" & lag(Result == "L", default = FALSE))
  ) %>%
  ungroup()

# Prepare modeling data
model_data <- all_results %>%
  mutate(
    win = ifelse(Result == "W", 1, 0),
    is_home = ifelse(Venue == "Home", 1, 0)
  ) %>%
  select(win, is_home, win_streak, lose_streak)
```

In this section, we fit a linear regression model to investigate how a team’s winning probability (win) is influenced by the following predictors:
- is_home: A binary indicator of whether the team is playing at home (1 = home, 0 = away)
- win_streak: The number of consecutive wins the team has before the current match.
- lose_streak: The number of consecutive losses the team has before the current match.

Although win is a binary outcome (0/1), we are here treating it with a linear model to get a quick and interpretable assessment of how these factors correlate with the team’s chance of winning. This model allows us to estimate a continuous outcome (predicted win probability) on a 0–1 scale, even though a logistic regression might be more common for binary data. Nonetheless, the linear approach is acceptable for exploratory analysis or illustrative purposes.

Our linear model is of the form:
$$
\hat{y} = \beta_0 
+ \beta_1 \cdot \texttt{is_home} 
+ \beta_2 \cdot \texttt{win_streak} 
+ \beta_3 \cdot \texttt{lose_streak},
$$

where:

- $\hat{y}$ s the predicted value of win (interpreted here as a proxy for the probability of winning).
- $\beta_0$ is the intercept (baseline outcome when all predictors are zero).
- $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients for the home venue, the current win streak, and the current loss streak, respectively.

```{r}
# Fit linear regression model
model <- lm(win ~ is_home + win_streak + lose_streak, data = model_data)


# Print results
print(summary(model))
```
#### Model Summary Interpretation
- **Intercept** ($\beta_0$ ≈ 0.299387)
  * This indicates the model’s baseline prediction when is_home = 0, win_streak = 0, and lose_streak = 0. A value of 0.299 suggests that, in the absence of other effects, a team has a predicted win rate (or probability) of about 29.9%.
- **is_home** ($\beta_1$ ≈ 0.131108)
  * A positive coefficient indicates that playing at home is associated with a higher likelihood of winning. Holding all else constant, being the home team increases the predicted win value by approximately 0.131. This effect is both positive and statistically significant (𝑝 < 0.001).
- **win_streak** ($\beta_2$ ≈ 0.018663)
  * This coefficient suggests that for each additional game in the current winning streak, the predicted win probability increases by about 0.019. This effect is also highly significant (𝑝 < 0.001), which aligns with the idea that teams on a winning streak tend to continue winning.
- **lose_streak** ($\beta_3$ ≈ -0.014855)  
  * A negative coefficient implies that for each additional game in the current losing streak, the predicted win probability decreases by about 0.015. This is likewise statistically significant (𝑝 < 0.001) and suggests that teams on a losing streak are somewhat less likely to win their next match.

**Model Fit:**
- **Multiple R-squared** = 0.1463 (Adjusted R-squared: 0.1446)
  * About 14.6% of the variation in the win variable is explained by these three predictors. While this indicates that the model leaves a good deal of variation unexplained (which is common in sports outcomes), the predictors do have a meaningful, statistically significant relationship with winning.


```{r}
# Calculate correlation matrix
correlation_matrix <- cor(model_data)

# Print correlation matrix
print(correlation_matrix)
```

The correlation matrix shows the pairwise linear relationships between win, is_home, win_streak, and lose_streak:

- **win and is_home:** A moderate positive correlation (0.14) implies that teams playing at home tend to win slightly more often.
- **win and win_streak:** A correlation of 0.28 indicates a positive relationship; the longer the current winning streak, the higher the chance of winning again.
- **win and lose_streak:** A correlation of about -0.19 indicates a negative relationship; the longer the current losing streak, the lower the chance of winning.
- **is_home with win_streak and lose_streak:** These correlations are near zero, suggesting that playing at home is relatively independent of a team’s current winning or losing streak in this dataset.

### Conclusion of the first Analysis
From this linear model, we see that venue (is_home), win streak, and lose streak each play an important role in predicting the likelihood of winning a match. The positive and negative coefficients for these predictors make intuitive sense:
- **Home Advantage** – Slight increase in win probability.
- **Winning Momentum** – Each consecutive win boosts chances for the next match.
- **Losing Momentum** – Each consecutive loss decreases chances for the next match.

Although the model explains only a modest portion of the variance in outcomes, it provides useful insights into how these factors correlate with a team’s probability of winning.

#### Visualization
```{r}
# Create visualization plots
venue_plot <- ggplot(model_data, aes(x = factor(is_home), y = win)) +
  geom_bar(stat = "summary", fun = "mean", fill = "skyblue") +
  geom_errorbar(stat = "summary", fun.data = "mean_se", width = 0.2) +
  scale_x_discrete(labels = c("Away", "Home")) +
  labs(x = "Venue", y = "Win Probability", title = "Win Probability by Venue") +
  theme_minimal()

win_streak_plot <- ggplot(model_data, aes(x = win_streak, y = win)) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  geom_jitter(alpha = 0.1, height = 0.05) +
  labs(x = "Win Streak", y = "Win Probability", title = "Win Probability by Win Streak") +
  theme_minimal()

lose_streak_plot <- ggplot(model_data, aes(x = lose_streak, y = win)) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  geom_jitter(alpha = 0.1, height = 0.05) +
  labs(x = "Lose Streak", y = "Win Probability", title = "Win Probability by Lose Streak") +
  theme_minimal()

coef_data <- data.frame(
  variable = names(coef(model))[-1],
  coefficient = coef(model)[-1],
  se = summary(model)$coefficients[-1, "Std. Error"]
)

coef_plot <- ggplot(coef_data, aes(x = variable, y = coefficient)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  geom_errorbar(aes(ymin = coefficient - 1.96 * se, 
                    ymax = coefficient + 1.96 * se), 
                width = 0.2) +
  labs(x = "Variables", y = "Coefficient Value", 
       title = "Model Coefficients with 95% CI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine and display all plots
combined_plots <- grid.arrange(
  venue_plot, win_streak_plot, 
  lose_streak_plot, coef_plot,
  ncol = 2
)

# Print results
print(combined_plots)
```


### Analysis of the home advantage over all seasons

```{r}
data_summary <- data %>%
  group_by(season) %>%
  summarise(
    total_matches = n(),
    home_wins     = sum(result == "H"),
    draws         = sum(result == "D"),
    away_wins     = sum(result == "A"),
    # Calculate average home attendance (igore NA-values)
    avg_home_attendance = mean(home_average_attendance, na.rm = TRUE)
  ) %>%
  mutate(
    pct_home_wins = 100 * home_wins / total_matches,
    pct_draws     = 100 * draws / total_matches,
    pct_away_wins = 100 * away_wins / total_matches
  )

data_summary

# 1) Keep only the relevant columns and reshape the data to long format
data_summary_long <- data_summary %>%
  select(season, pct_home_wins, pct_draws, pct_away_wins) %>%
  pivot_longer(
    cols = c(pct_home_wins, pct_draws, pct_away_wins),
    names_to = "outcome",
    values_to = "pct"
  )

# 2) Plot as a line chart
ggplot(data_summary_long, aes(x = season, y = pct, color = outcome, group = outcome)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Season",
    y = "Percentage",
    color = "Outcome",
    title = "Distribution of Match Outcomes by Season"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # tilt labels if needed
  )


```
Interpretation:
We clearly see that the home team wins far more games then the away team. This counts for every season. With this analysis, we lay the basis that the home advantage is indeed correct and we can go further and try to find out, if the home advantage gets bigger, the bigger the attendance is for a game.


## Grouped analysis by attendancies
```{r}
data %>%
  mutate(
    attendance_group = case_when(
      home_average_attendance < 25000 ~ "<25k",
      home_average_attendance < 35000 ~ "25k-35k",
      home_average_attendance < 50000 ~ "35k-50k",
      TRUE                            ~ ">50k"
    )
  ) %>%
  group_by(attendance_group) %>%
  summarise(
    total_matches = n(),
    avg_goal_diff = mean(home_goals - away_goals, na.rm = TRUE),
    pct_home_wins = mean(result == "H", na.rm = TRUE) * 100
  )


# Create 'attendance_group', reorder its levels, summarize, and store results in 'data_summary'
data_summary <- data %>%
  mutate(
    attendance_group = case_when(
      home_average_attendance < 25000 ~ "<25k",
      home_average_attendance < 35000 ~ "25k-35k",
      home_average_attendance < 50000 ~ "35k-50k",
      TRUE                            ~ ">50k"
    )
  ) %>%
  # Reorder factor levels so they appear in the desired sequence
  mutate(
    attendance_group = factor(
      attendance_group,
      levels = c("<25k", "25k-35k", "35k-50k", ">50k")
    )
  ) %>%
  # Summarize: count total matches, home wins, and compute home-win percentage
  group_by(attendance_group) %>%
  summarise(
    total_matches = n(),
    home_wins     = sum(result == "H", na.rm = TRUE),
    pct_home_wins = 100 * home_wins / total_matches
  )

# Plot a bar chart of home-win percentage by attendance group
ggplot(data_summary, aes(x = attendance_group, y = pct_home_wins)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Attendance Group",
    y = "Home Win Percentage",
    title = "Home Win Percentage by Attendance Group"
  ) +
  theme_minimal()


```
Interpretation:
We grouped the attendances for all the games into these four groupes and calculated the mean home_game_win_percentage for every group. Clearly, we see a trend, that the win percentage gets bigger, with more people in the stadium. But we have to be careful, because we have a confounder. The confounder is that the best teams (this counts for most if not every country) have the biggest stadiums (bigger fanbase etc.). Therefore its no surprise that the play in front of bigger crowds and do also win more games at home. 

## Correlation between goals/points with attendance
```{r}
# Correlation between goals with attendance
cor(data$home_goals - data$away_goals, data$home_average_attendance, use = "complete.obs")

# Correlation between points with attendance
cor(data$points_home, data$home_average_attendance, use = "complete.obs")

```
Interpretation:
We calculated a correlation between goals with attendance of 0.26 and points with attendance of 0.23. Therefore we have a small correlation between those three variables.

## look for teams whos attendance changed drastically
```{r}
df_changes <- data %>%
  # 1) Group by team
  group_by(home_team) %>%
  # 2) Sort by season within each team
  arrange(season, .by_group = TRUE) %>%
  # 3) Compute attendance difference & percentage difference
  mutate(
    attendance_diff = home_average_attendance - lag(home_average_attendance),
    attendance_diff_pct = (attendance_diff / lag(home_average_attendance)) * 100
  ) %>%
  # 4) Keep only rows with large (positive or negative) changes
  filter(!is.na(attendance_diff)) %>%  # remove the first row per team (no lag)
  filter(abs(attendance_diff_pct) > 15)

# View the teams/seasons with “drastic” changes
df_changes

```

Interpretation:
As we have a confounder, we looked through our dataset to find teams whos season home attendance changed drastically to the next season. We found 4 teams/season in which the attendance changed at least 15% (more or less). Then we did a difference-in-difference analysis, to see whether the bigger/smaller attendance could have an impact on the goal-difference of the home team.

## Difference in difference
### Tottenham Hotspurs
```{r}
data_spurs <- data %>%
  filter(season %in% c("2015-2016", "2016-2017", "2017-2018"))

data_spurs <- data_spurs %>%
  mutate(
    treat = if_else(home_team == "Tottenham Hotspur" | away_team == "Tottenham Hotspur", 1, 0),
    post  = if_else(season == "2017-2018", 1, 0)
  )

did_mod <- lm((home_goals - away_goals) ~ treat + post + treat:post, data = data_spurs)
summary(did_mod)


```

### Liverpool
```{r}
data_liv <- data %>%
  # Keep only the four relevant seasons
  filter(season %in% c("2014-2015", "2015-2016", "2016-2017", "2017-2018")) %>%
  
  # Mark Liverpool as treated (either as home team or away team if you want both),
  # but here we'll mirror the earlier approach and only mark "home_team" == "Liverpool".
  mutate(
    treat = if_else(home_team == "Liverpool", 1, 0),
    
    # post = 1 for 2016-17 and 2017-18 (the period after the attendance jump),
    # 0 for 2014-15 and 2015-16
    post = if_else(season %in% c("2016-2017", "2017-2018"), 1, 0),
    
    # Our outcome variable: goal difference from the home team's perspective
    goal_diff = home_goals - away_goals
  )

did_mod_liv <- lm(goal_diff ~ treat + post + treat:post, data = data_liv)
summary(did_mod_liv)

```

### Man City

```{r}
data_mc <- data %>%
  # Keep only the two seasons of interest
  filter(season %in% c("2014-2015", "2015-2016")) %>%
  mutate(
    treat = if_else(home_team == "Manchester City", 1, 0),
    post  = if_else(season == "2015-2016", 1, 0),
    goal_diff = home_goals - away_goals
  )

did_mod_mc <- lm(goal_diff ~ treat + post + treat:post, data = data_mc)
summary(did_mod_mc)

```


### West Ham

```{r}
data_wh <- data %>%
  # 1) Filter for the relevant seasons
  filter(season %in% c("2014-2015", "2015-2016", "2016-2017", "2017-2018")) %>%
  
  # 2) Create treat, post, and the outcome variable
  mutate(
    treat = if_else(home_team == "West Ham United", 1, 0),
    post  = if_else(season %in% c("2016-2017", "2017-2018"), 1, 0),
    goal_diff = home_goals - away_goals
  )

did_mod_wh <- lm(goal_diff ~ treat + post + treat:post, data = data_wh)
summary(did_mod_wh)

```
Interpretation of difference-in-difference:
For two teams we found a significant impact on attendance on the average goal-difference and for the other two teams we did not. Therefore its hard to say, wether we can clearly state that the bigger home crowd gets those teams an advantage or not.


## Attendancies for these 4 teams
```{r}
# Define the teams and seasons of interest
teams_of_interest <- c("Tottenham Hotspur", "Liverpool", 
                       "Manchester City", "West Ham United")
seasons_of_interest <- c("2013-2014", "2014-2015", "2015-2016", 
                         "2016-2017", "2017-2018")

# 1) Filter to the teams & seasons you want
# 2) Group by season & home_team
# 3) Summarize average attendance (ignoring NA values)
# 4) Pivot wider so each team is a separate column
# 5) Arrange by season for a nice chronological table
attendance_summary <- data %>%
  filter(home_team %in% teams_of_interest,
         season %in% seasons_of_interest) %>%
  group_by(season, home_team) %>%
  summarise(avg_home_attendance = mean(home_average_attendance, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = home_team, values_from = avg_home_attendance) %>%
  arrange(season)

attendance_summary

```

Interpretation:
Here I just wanted to have a table, which I can copy on the presentation, for which we see how and when the attendance changes.



